A Tutorial on Tailored Simulation-Based Power Analysis for Experimental Designs with Generalized Linear Mixed Models

### 2023-10-19 docs/manuscript.Rmd
removed hci references from manuscript

  When planning experimental research, determining an appropriate sample size and using suitable statistical models are crucial for robust and informative results. However, the recent replication crisis in Human-Computer Interaction (HCI) and other empirical research fields underlines the need for more rigorous statistical methodology and well-powered designs. Generalized linear mixed models (GLMMs) offer a flexible statistical framework to analyze experimental data with complex (e.g., dependent and hierarchical) data structures. Yet, analytic methods and software cannot be applied to conduct a priori power analyses for GLMMs, necessitating data simulation approaches. Based on a practical case study, the current tutorial equips researchers with a step-by-step guide and corresponding code for conducting tailored a priori power analyses to determine appropriate sample sizes with GLMMs. Finally, we give an outlook on the increasing importance of simulation-based power analysis in experimental research.

@@

<!-- General short intro paragraph, need for well-powered and analyzed experimental research -->
When planning experimental research, it is essential to determine an appropriate sample size to ensure that the results obtained are both robust and informative, and to use appropriate statistical models to analyze the data [@lakensSampleSizeJustification2022]. However, the recent replication crisis in Human-Computer Interaction (HCI) and several other disciplines grounded on empirical research has illustrated many challenges surrounding the reproducibility and reliability of findings [@cockburnThreatsReplicationCrisis2020; @robertsonImprovingStatisticalPractice2016; @yarkoniGeneralizabilityCrisis2022]. As a result, there is a growing need for more rigorous statistical methodology and the adoption of well-powered experimental designs. While software solutions exist for simple statistical models and experimental designs, many researchers lack the skills and tools to conduct "a priori" (i.e., before data collection) power analyses for more complex research designs using the flexible generalized linear mixed models (GLMM) framework in order to determine the required sample size in their experiments. In the present work, we provide a tutorial consisting of a concrete example for tailored a priori power analyses using data simulations based on GLMMs.

@@

<!-- Advantages of GLMMs over "simpler" models -->
As study designs become more complex, researchers require more sophisticated statistical models to capture the nuanced relationships and grouping structures introduced by their study designs [@yarkoniGeneralizabilityCrisis2022]. GLMMs (also called multilevel models) are gaining increasing popularity in analyzing data in HCI and other empirical disciplines because they offer a flexible framework for handling data with outcome variables that are not normally distributed (e.g., categorical outcomes) while accounting for both fixed and random effects [@kapteinUsingGeneralizedLinear2016; @fahrmeirRegressionModelsMethods2021].

>>>>

  When planning experimental research, determining an appropriate sample size and using suitable statistical models are crucial for robust and informative results. However, the recent replication crisis underlines the need for more rigorous statistical methodology and well-powered designs. Generalized linear mixed models (GLMMs) offer a flexible statistical framework to analyze experimental data with complex (e.g., dependent and hierarchical) data structures. Yet, analytic methods and software cannot be applied to conduct a priori power analyses for GLMMs, necessitating data simulation approaches. Based on a practical case study, the current tutorial equips researchers with a step-by-step guide and corresponding code for conducting tailored a priori power analyses to determine appropriate sample sizes with GLMMs. Finally, we give an outlook on the increasing importance of simulation-based power analysis in experimental research.

@@

<!-- General short intro paragraph, need for well-powered and analyzed experimental research -->
When planning experimental research, it is essential to determine an appropriate sample size to ensure that the results obtained are both robust and informative, and to use appropriate statistical models to analyze the data [@lakensSampleSizeJustification2022]. However, the recent replication crisis has illustrated many challenges surrounding the reproducibility and reliability of findings [@yarkoniGeneralizabilityCrisis2022]. As a result, there is a growing need for more rigorous statistical methodology and the adoption of well-powered experimental designs. While software solutions exist for simple statistical models and experimental designs, many researchers lack the skills and tools to conduct "a priori" (i.e., before data collection) power analyses for more complex research designs using the flexible generalized linear mixed models (GLMM) framework in order to determine the required sample size in their experiments. In the present work, we provide a tutorial consisting of a concrete example for tailored a priori power analyses using data simulations based on GLMMs.

@@

<!-- Advantages of GLMMs over "simpler" models -->
As study designs become more complex, researchers require more sophisticated statistical models to capture the nuanced relationships and grouping structures introduced by their study designs [@yarkoniGeneralizabilityCrisis2022]. GLMMs (also called multilevel models) are gaining increasing popularity in analyzing data because they offer a flexible framework for handling data with outcome variables that are not normally distributed (e.g., categorical outcomes) while accounting for both fixed and random effects [@fahrmeirRegressionModelsMethods2021].
